Fail Faster

### 2025-03-19 paper/paper.tex
last bit of intro work

  To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
 First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs ``without regret''~\cn.
We apply this technique to generators, completely eliminating the run-time overhead of
many common generator abstractions.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \iflater\hg{More words here for
parallelism?}\fi Together, these insights provide a recipe for improving existing generator libraries.

To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} --- a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street ---
and ScalaCheck --- a very mature PBT library and the de facto choice in Scala.
In both languages, our staged versions are direct drop-in replacements that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, the same generator written once with \name{}
and once with the original generator library produces exactly the same sequence of values.
The semantic equivalence between the original generators and \name generators allows us to 
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

We evaluate the optimizations in \name with a series of case studies in each
generator library, and show that the \name generators run faster than their unstaged equivalence.
Further, we show that this performance benefit\hg{wc?} translates to bug-finding
ability by running these case studies in the Etna framework~\cite{etna}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.

In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiencies in PBT generator libraries, which can significantly impact performance: 
    namely, and abstraction overhead, and \rand choice. \secref{section:motiv}
    % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
    \item We implement our approach in OCaml and Scala, showcasing its cross-language applicability.
    \secref{section:4}
    \item We construct generators in \name and \scalaname, and show that they give substantial performance improvements. Finally, we show
    that these performance improvements extend to bug-finding ability.
     \secref{section:eval}
\end{enumerate}

\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}\todo{Todo Thia}

@@

\section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
\jwc{This section needs a much better title...}
\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}

>>>>

To address this question, we outline two key principles for designing generator libraries that balance efficiency with flexibility.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used to
build DSLs ``without regret''~\cn{}.
We apply this technique to generators, completely eliminating the run-time overhead of
many common generator abstractions.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. Together, these insights provide a recipe for improving existing generator libraries.

To demonstrate the effectiveness of our approach, we apply it to two PBT generator libraries, namely \bq{} \cn{} --- a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street ---
and ScalaCheck \cn{} --- a very mature PBT library and the de facto choice in Scala.
In both languages, our staged versions are direct drop-in replacements that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, the same generator written once with \name{}
and once with the original generator library produces exactly the same sequence of values.
The semantic equivalence between the original generators and \name generators allows us to 
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.

We evaluate the optimizations in \name with a series of case studies in each
generator library, and show that the \name generators run faster than their unstaged equivalence.
Further, we show that this performance benefit\hg{wc?} translates to bug-finding
ability \jwc{The paper is called ``fail faster'' --- we shouldn't say ``bug finding ability'', we should say something like ``how fast you can squash bugs''. This is too passive }by running these case studies in the Etna framework~\cite{etna}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.

In summary, we show that monadic PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions.
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance: 
    namely and abstraction overhead and \rand choice. 
    % \item We apply multi-stage programming to the novel domain of generators. \jwc{Less sure this should be a top-level conrib.} \secref{section:}
    \item We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that both \name{} and choosing a faster \rand yield substantial performance improvements. Finally, we show
    that these performance improvements extend to significantly improved bug-finding ability. \jwc{This should be sell-ier: 2x speedups, etc. ``bug-finding ability'' isn't a race against time.}
\end{enumerate}

\jwc{Ensure we have a signposting table-of-contents-ish few sentences here, if it's not covered by the contribs.}\todo{Todo Thia}
In Section~\ref{section:motiv}, we introduce monadic generator DSLs, discuss how they work, and take an example-guided tour through some of the ways that
they are slower than is possible.
Then, in Section~\ref{section:faster-generators}, we explain multi-stage programming,
iteratively develop the \name{} technique, and discuss options for faster \rand{} \jwc{fix}.
Section~\ref{section:eval} evaluates all of the previous work, demonstrating the generator performance improvements
and how they lead to much faster bug finding.

@@

\section{Speeding up Generator DSLs by Staging and Faster Randomness Libraries}
\label{section:faster-generators}
\jwc{This section needs a much better title...}
\jwc{Emphasize that this is an \emph{COMPLETELY EQUIVALENT drop in replacement!} We din't just build a different library with different distributions.}
